---
title: "DS-University-Enrolment"
author: "Matthew Adenyo"
date: "2023-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

You are working as a data scientist at a local University. The university started offering online
courses to reach a wider range of students. The university wants you to help them understand
enrollment trends. They would like you to identify what contributes to higher enrollment. In
particular, whether the course type (online or classroom) is a factor.

The dataset contains records for each course offered over the last 5 years.


```{r library}
#Load required libraries
library(readr)
library(ggplot2)
library(caret)
library(dplyr)
library(tidyr)
library(glmnet)
library(lattice)
library(e1071)
```

```{r data}
#Read datasets
data <- read.csv("D:/Data Science Projects/DS-Associate-Certification/Data/university_enrollment_2306.csv")
```


```{r summary}
# Display data summary
str(data)
dim(data)
sapply(data, class)
summary(data)
```


# Task 1:
## For every column in the data:
### a. State whether the values match the description given in the table above.
### b. State the number of missing values in the column.
### c. Describe what you did to make values match the description if they did not match.



```{r}
# Compare the description for each column

# course_id Data Validation
unique_id <- length(unique(data$course_id)) == nrow(data)
course_count_na <- sum(is.na(data$course_id))

cat("----- course_id -----\n")
cat(paste("Column 'course_id' has only unique values:", ifelse(unique_id, "TRUE", "FALSE"), "\n"))
cat(paste("Sum of missing values:", course_count_na, "\n\n"))

# course_type Data Validation
online_class <- unique(data$course_type)
course_type_count_na <- sum(is.na(data$course_type))

cat("----- course_type -----\n")
cat("Values in this column:\n", online_class, "\n")
cat(paste("Sum of missing values:", course_type_count_na, "\n\n"))

# year Data Validation
years <- unique(data$year)
year_count_na <- sum(is.na(data$year))

cat("----- year -----\n")
cat("Values in this column:\n", years, "\n")
cat(paste("Sum of missing values:", year_count_na, "\n\n"))

# enrollment_count Data Validation
enrollment_count_na <- sum(is.na(data$enrollment_count))

cat("----- enrollment_count -----\n")
cat("Description of values in this column:\n")
summary(data$enrollment_count)
cat(paste("Sum of missing values:", enrollment_count_na, "\n\n"))

# pre_score Data Validation
# Replace non-numeric values with NA
data$pre_score[data$pre_score == "-"] <- NA
pre_score_count_na <- sum(is.na(data$pre_score))

cat("----- pre_score -----\n")
cat("Description of values in this column:\n")
summary(as.numeric(data$pre_score))
cat(paste("Sum of missing values before cleaning:", pre_score_count_na, "\n\n"))

# post_score Data Validation
post_score_count_na <- sum(is.na(data$post_score))

cat("----- post_score -----\n")
cat("Description of values in this column:\n")
summary(data$post_score)
cat(paste("Sum of missing values before cleaning:", post_score_count_na, "\n\n"))

# pre_requirement Data Validation
pre_req <- unique(data$pre_requirement)
pre_req_count_na <- sum(is.na(data$pre_requirement))

cat("----- pre_requirement -----\n")
cat("Values in this column:", pre_req, "\n")
cat(paste("Sum of missing values before cleaning:", pre_req_count_na, "\n\n"))

# department Data Validation
dep <- unique(data$department)
dep_count_na <- sum(is.na(data$department))

cat("----- department -----\n")
cat("Values in this column:\n", dep, "\n")
cat(paste("Sum of missing values before cleaning:", dep_count_na, "\n\n"))

```



```{r}
# Data Cleaning
cat("For each column that presented missing values, or incorrect description:\n")
# pre_score - replace missing values and "-"
data$pre_score[is.na(data$pre_score) | data$pre_score == "-"] <- 0
pre_score_count_na <- sum(is.na(data$pre_score))
cat("----- pre_score -----\n")
cat(paste("Sum of missing values after cleaning:", pre_score_count_na, "\n\n"))

# post_score - replace missing values and "-"
data$post_score[is.na(data$post_score) | data$post_score == "-"] <- 0
post_score_count_na <- sum(is.na(data$post_score))
cat("----- post_score -----\n")
cat(paste("Sum of missing values after cleaning:", post_score_count_na, "\n\n"))

# pre_requirement - replace missing values and "-"
data$pre_requirement[is.na(data$pre_requirement) | data$pre_requirement == "-"] <- "None"
pre_req_count_na <- sum(is.na(data$pre_requirement))
cat("----- pre_requirement -----\n")
cat(paste("Sum of missing values after cleaning:", pre_req_count_na), "\n")
cat(paste("Unique values after cleaning:", unique(data$pre_requirement), "\n\n"))

# department - replace "Math" values with "Mathematics"
data$department <- ifelse(data$department == "Math", "Mathematics", data$department)
cat("----- department -----\n")
cat(paste("Unique values after cleaning:", unique(data$department), "\n"))
```


**Solution To Task 1**

The dataset contains **1850 rows and 8 columns** with missing values before cleaning. All the columns were validated against the criteria in the dataset table. Below were the observations:


* Column 1 (course_id): Matches the description as unique identifier with 0 missing values.
* Column 2 (course_type): All values are either "online" or "classroom" with 0 missing values.
* Column 3 (year): All values are between 2011 and 2022 inclusive with 0 missing values.
* Column 4 (enrollment_count): All values are discrete counts of enrollments with 0 missing values.
* Column 5 (pre_score): Contains continuous score data with **130 missing values.** Missing values were replaced with 0.
* Column 6 (post_score): Contains continuous score data with **185 missing values.** Missing values were replaced with 0.
* Column 7 (pre_requirement): All values are either "None", "Beginner" or "Intermediate" with **89 missing values.** Missing values were replaced with "None".
* Column 8 (department): The values here are either "Science", "Technology", "Engineering", or "Mathematics", however, there were instances of *"Math"* which is not in the given criteria. It is assumed that "Math" refers to "Mathematics" and were replaced accordingly.



# Task 2:
## Describe the distribution of the enrollment counts. Your answer must include a visualization that shows the distribution.

```{r}
# Creating subplots
par(mfrow=c(1, 2), mar=c(4, 4, 2, 1))

# Distribution of Enrollment Count
ggplot(data, aes(x = enrollment_count)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
 # geom_density(alpha = 0.5, color = "red") +
  labs(title = "Distribution of Enrollment Counts") +
  theme_minimal()


# Boxplot of Enrollment Count
ggplot(data, aes(y = enrollment_count)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Boxplot of Enrollment Counts") +
  theme_minimal()

# Reset the plotting layout
par(mfrow=c(1, 1))

# Calculate important metrics
mean_val <- mean(data$enrollment_count)
median_val <- median(data$enrollment_count)
mode_val <- as.numeric(names(table(data$enrollment_count))[table(data$enrollment_count)
                                                           == max(table(data$enrollment_count))])

cat(paste("- Mean:", mean_val, "\n"))
cat(paste("- Median:", median_val, "\n"))
cat(paste("- Mode:", mode_val, "\n"))

```



**Solution To Task 2**



```{r}

#TASK 3
#Bar plot to Visualize the Course type
ggplot(data, aes(x = course_type, fill = course_type)) +
  geom_bar() +
  labs(title = "Number of Courses by Type") +
  theme_minimal() +
  theme(legend.position = "none")

# Calculate and print the proportion
classroom <- sum(data$course_type == "classroom")
online <- nrow(data) - classroom
proportion <- round((online / (online + classroom)) * 100)

cat(paste("- Proportion of Online to Classroom Type of Courses:", proportion, "%\n"))


```



```{r}

#TASK 4:
# Violin plot
ggplot(data, aes(x = course_type, y = enrollment_count)) +
  geom_violin(fill = "skyblue", color = "black", alpha = 0.5) +
  labs(title = "Enrollment Count by Course Type") +
  theme_minimal()

# Calculate and print the means
classroom <- round(mean(data[data$course_type == "classroom", "enrollment_count"], na.rm = TRUE))
online <- round(mean(data[data$course_type == "online", "enrollment_count"], na.rm = TRUE))

cat(paste("- Avg. enrollment count for Classroom Courses:", format(classroom, big.mark = ","), "\n"))
cat(paste("- Avg. enrollment count for Online Courses:", format(online, big.mark = ","), "\n"))

```



```{r}

#TASK 5:
#converting all categorical variables to factors and apply OneHotEncoder

# Convert 'pre_score' to a factor variable
data$pre_score <- as.factor(data$pre_score)

# Apply one-hot encoding to categorical variables
data <- data %>%
  dplyr::mutate(across(pre_score, ~ as.integer(.)))

# Splitting the data into training and testing sets
set.seed(123)  # For reproducibility
splitIndex <- createDataPartition(data$enrollment_count, p = 0.8, list = FALSE)
train_data <- data[splitIndex, ]
test_data <- data[-splitIndex, ]

```




```{r}

#TASK 6:
# Fit a Linear Regression model (LM)
lm_model <- lm(enrollment_count ~ ., data = train_data)

# Make predictions using the LM model
predictions <- predict(lm_model, newdata = test_data)

# Calculate Root Mean Squared Error (RMSE)
rmse <- sqrt(mean((predictions - test_data$enrollment_count)^2))
cat("Root Mean Squared Error (LM):", rmse, "\n")

# Create a data frame for plotting
plot_data <- data.frame(
  actual = test_data$enrollment_count,
  predicted = predictions
)

# Create a scatter plot comparing actual and LM predicted values
ggplot(plot_data, aes(x = actual, y = predicted)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(x = "Actual Enrollment", y = "LM Predicted Enrollment",
       title = "Actual vs. LM Predicted Enrollment") +
  annotate("text", x = max(plot_data$actual), y = min(plot_data$predicted),
           label = paste("RMSE =", round(rmse, 2)), hjust = 1)

```



```{r}

#TASK 7:
# Fit a Support Vector Machine (SVM) model
svm_model <- svm(enrollment_count ~ ., data = train_data)

# Make predictions using the SVM model
predictions <- predict(svm_model, newdata = test_data)

# Calculate Root Mean Squared Error (RMSE)
rmse <- sqrt(mean((predictions - test_data$enrollment_count)^2))
cat("Root Mean Squared Error (SVM):", rmse, "\n")

# Create a data frame for plotting
plot_data <- data.frame(
  actual = test_data$enrollment_count,
  predicted = predictions
)

# Create a scatter plot comparing actual and SVM predicted values
ggplot(plot_data, aes(x = actual, y = predicted)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(x = "Actual Enrollment", y = "SVM Predicted Enrollment",
       title = "Actual vs. SVM Predicted Enrollment") +
  annotate("text", x = max(plot_data$actual), y = min(plot_data$predicted),
           label = paste("RMSE =", round(rmse, 2)), hjust = 1)

```





Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
